---
title: "R Notebook"
output: html_notebook
---

Building out a notebook for my notes - seems legit.

Ran across a paper 'Content-boosted Matrix Factorization Techniques for Recommender Systems'
and realized 2 things:

1) These methods could be pretty useful in my work

2) I do not know enough matrix math

To address point 2 - I am going to work through the math using R as my 
terra firma and see if I can build a new package OR extend an existing one to 
do a service to the open source community - plus L E A R N I N G!

##SVD

SVD - or singular value decomposition - allows you take a matrix and create two 
new matricies that have compressed information (and thus size) via a user specified 
parameter.  More formally - a matrix M gets converted to
U S V where U is the number of rows in M x the user specified decomposition term
and V is the number of columns x the user specified decomposition term.  S is 
the rectangular diagonal matrix that allows you to get U and V back to M's dimensions.

```{r}
test_matrix = matrix(data = c(c(1,1,1,0,0),
                              c(1,1,0,0,0),
                              c(1,1,1,1,1),
                              c(0,0,0,0,0),
                              c(0,0,0,1,1)),
                     ncol = 5)

num_users = nrow(test_matrix)
num_movies = ncol(test_matrix)

num_features = 2
maxit = 100
                

eigen(test_matrix)

A   <- matrix(c(3,1,1,3),2,2)
eA  <- eigen(A) 
```

```{r}
A = matrix(data = c(c(2,1,0,0),
                    c(4,3,0,0),
                    c(0,0,1,1),
                    c(5,0,1,0),
                    c(2,0,2,2),
                    c(0,3,0,2)),
                     ncol = 6)


A_svd <- svd(A)
A_svd

AtA = t(A) %*% A

AtA_eigen = eigen(AtA)

V_component = AtA_eigen$vectors
V_eigen_values = round(AtA_eigen$values, 7)
V_component = AtA_eigen$vectors[,V_eigen_values > 0]
AAT = A %*% t(A)

AAT_eigen = eigen(AAT)

U_component = AAT_eigen$vectors
#Only interestin in the non-xero eigenvalues and their respective eigenvectors
U_eigen_values = round(AAT_eigen$values, 7)
U_component = U_component[,U_eigen_values > 0]

r = sqrt(U_eigen_values)

dimensions = 2
# r_num = length(r)
r_reduced = diag(r[1:dimensions])
U_component_reduced = U_component[,1:dimensions]
V_component_reduced = V_component[,1:dimensions]

U_component_reduced %*% r_reduced %*% t(V_component_reduced)


U_component %*% r %*% t(V_component)


ds <- diag(A_svd$d[1:2])  # let's just use the first two values
us <- as.matrix(A_svd$u[, 1:2])
vs <- as.matrix(A_svd$v[, 1:2])
m.approx1 <- us %*% ds %*% t(vs)
m.approx1


ATA
num_users = nrow(test_matrix)
num_movies = ncol(test_matrix)

num_features = 2
maxit = 100
                

AtA = eigen(test_matrix%*%t(test_matrix))

tAA = eigen(t(test_matrix)%*%(test_matrix))

apply(AtA$vectors, 2, sqrt)
sqrt(eigen(t(test_matrix)%*%(test_matrix)))

A   <- matrix(c(3,1,1,3),2,2)
eA  <- eigen(A) 
```

```{r}
attribute_matrix = matrix(data = c(c(1,1,0,0),
                                      c(1,1,0,0),
                                      c(0,0,1,1),
                                      c(1,0,0,1)),
                             ncol = 4
)
c = 2
V_component = A_svd$v
qtq = t(V_component[,1]) %*% V_component[,4]
SC_i = t(attribute_matrix[,1]) %*% attribute_matrix[,4] >= c
qtq/SC_i
qtq
```

```{r}
a = c(1,0,1,0)
ai = c(1,1,1,0)
t(a)%*%ai 
```





```{r}

    x <- as.matrix(test_matrix)

    dx <- dim(x)
    n <- dx[1L]
    p <- dx[2L]

    nu = min(n, p)
    nv = min(n, p)
    # La.res <- La.svd(x, nu, nv)
    #La.svd#####################################################################
   
    x <- as.matrix(x)
    n <- nrow(x)
    p <- ncol(x)

    zero <- if (is.complex(x)){
      0 + (0+0i)
    }else 0
        
    
    if (nu || nv) {
        np <- min(n, p)
        if (nu <= np && nv <= np) {
            jobu <- "S"
            u <- matrix(zero, n, np)
            vt <- matrix(zero, np, p)
            nu0 <- nv0 <- np
        }
        else {
            jobu <- "A"
            u <- matrix(zero, n, n)
            vt <- matrix(zero, p, p)
            nu0 <- n
            nv0 <- p
        }
    }
    
    res <- if (is.complex(x)) 
        .Internal(La_svd_cmplx(jobu, x, double(min(n, p)), u, 
            vt))
    La.svd
    else .Internal(La_svd(jobu, x, double(min(n, p)), u, vt))
    res <- res[c("d", if (nu) "u", if (nv) "vt")]
    if (nu && nu < nu0) 
        res$u <- res$u[, seq_len(min(n, nu)), drop = FALSE]
    if (nv && nv < nv0) 
        res$vt <- res$vt[seq_len(min(p, nv)), , drop = FALSE]
    res
}
    res <- list(d = La.res$d)
    if (nu) 
        res$u <- La.res$u
    if (nv) {
        if (is.complex(x)) 
            res$v <- Conj(t(La.res$vt))
        else res$v <- t(La.res$vt)
    }
    res
```


```{r}
unroll_Vecs <- function (params, Y, R, num_users, num_movies, num_features) {
  # Unrolls vector into X and Theta
  # Also calculates difference between preduction and actual 
  
  endIdx <- num_movies * num_features
  
  X     <- matrix(params[1:endIdx], nrow = num_movies, ncol = num_features)
  Theta <- matrix(params[(endIdx + 1): (endIdx + (num_users * num_features))], 
                  nrow = num_users, ncol = num_features)
  
  Y_dash     <-   (((X %*% t(Theta)) - Y) * R) # Prediction error
  
  return(list(X = X, Theta = Theta, Y_dash = Y_dash))
}

J_cost <-  function(params, Y, R, num_users, num_movies, num_features, lambda, alpha) {
  # Calculates the cost
  
  unrolled <- unroll_Vecs(params, Y, R, num_users, num_movies, num_features)
  X <- unrolled$X
  Theta <- unrolled$Theta
  Y_dash <- unrolled$Y_dash
  
  J <-  .5 * sum(   Y_dash ^2)  + lambda/2 * sum(Theta^2) + lambda/2 * sum(X^2)
  
  return (J)
}

grr <- function(params, Y, R, num_users, num_movies, num_features, lambda, alpha) {
  # Calculates the gradient step
  # Here lambda is the regularization parameter
  # Alpha is the step size

  unrolled <- unroll_Vecs(params, Y, R, num_users, num_movies, num_features)
  X <- unrolled$X
  Theta <- unrolled$Theta
  Y_dash <- unrolled$Y_dash
  
  X_grad     <- ((   Y_dash  %*% Theta) + lambda * X     )
  Theta_grad <- (( t(Y_dash) %*% X)     + lambda * Theta )
  
  grad = c(X_grad, Theta_grad)
  return(grad)
}

# Now that everything is set up, call optim

res <- optim(par = c(runif(num_users * num_features), 
                     runif(num_movies * num_features)), # Random starting parameters
             fn = J_cost,
             gr = grr, 
             Y=Y, 
             R=R, 
             num_users=num_users, 
             num_movies=num_movies,
             num_features=num_features, 
             lambda=lambda, 
             alpha = alpha, 
             method = "L-BFGS-B",
             control=list(maxit=maxit, trace=1))

```

